---
title: "Growth-defense trade-off article analyses"
author: "Julia Harenčár"
date: "3/29/2023"
execute:
  echo: true #will keep code chunks
  warning: false #removes warning functions
format: 
  html: 
    self-contained: true #makes one html file to render that is sharable
    page-layout: full #uses the full page
    code-fold: true #lets user fold up codes
    code-tools: true #makes an option in the upper left for users to reveal/hide all code and copy source code
    #makes long lines of code rap
    code-overflow: wrap
    #makes table of contents
    toc: true
    #sets location of table of contents
    toc-location: left
#set theme settings. in this case user can switch between light and dark mode
theme:
  #give light option
  light: flatly
  #give dark option
  dark: darkly
---

Analyses of herbivory and chemistry data for the article, "Growth-defense trade-offs facilitate species coexistence: empirical evidence from recently diverged tropical plants."

Loading packages and setting defaults:

```{r}
#| warning: false
# analysis
library("DHARMa")
library("car")
library("lme4")
library("glmmTMB") 
# plotting
library("tidyverse")
library("ggplot2")
library("patchwork")
library("RColorBrewer")
theme_set(theme_bw())

```

## Wild herbivory

Analyses comparing wild herbivory between *Costus villosissimus* and *C. allenii*.

### Presence v absence

Analysis comparing the presence or absence of herbivory on wild plants from both 2019 and 2022.

First, we read in the 2019 and 2022 data, create presence/absence columns for each, and combine them into a single dataframe.

```{r}
#| warning: false
# 2019 data
Wherb19 <- read.csv("vill-all_RLB-herbivory_data_2019-Herbivory.csv", header=TRUE)
# create presence/absence of herbivory column
Wherb19 <- Wherb19 %>% mutate(pa = case_when(
  per_herbivory == 0 ~ 0,
  per_herbivory != 0 ~ 1
))

# 2020 data
Wherb22 <- read.csv("2022_wild_herbivory.csv", header = T)
# create presence/absence of herbivory column
Wherb22 <- Wherb22 %>% 
  mutate(pa = case_when(
  num_w_RLBH == 0 ~ 0,
  num_w_RLBH != 0 ~ 1
)) %>% dplyr::select(date, spp, id, pa)

# making combined 2019/2022 dataset
Wherb19$year <- '2019'
Wherb22$year <- '2022'
names(Wherb19) <- c("date","ID", "spp", "ave_per_herb", "pa", "year")
Wherb_full <- Wherb22 %>% 
  dplyr::select(spp,pa,year) %>% 
  full_join((Wherb19 %>% dplyr::select( "spp", "pa", "year")),.) 
```

Next, we evaluate the potential impacts of plant species and year on the presence vs absence of herbivory with a binomial regression. We first check to ensure the model is not under- or over-dispersed.

```{r}
# binomial regression 
bin_reg <- glm(Wherb_full$pa ~ Wherb_full$spp + Wherb_full$year, family = "binomial")

# Check over or under-dispersion with the package DHARMA. 
simOut <- simulateResiduals(bin_reg)
plot(simOut) # dispersion looks good
testDispersion(simOut) # dispersion = 1.0241, p-value = 0.808; not overdispersed
```

The model is not overdispersed! :D

Now we look at the results of the regression and plot wild herbivory by year and species.

```{r}
# binomial regression summary
summary(bin_reg)

# # code for choosing colorblind friendly colors. 
# library(colorblindcheck)
# palette_check(c("#cc79a7", "#f0e442"), plot = TRUE)

# Making a simple barplot of presence/absence herbivory data from both years
# calculate percentages for plot
PA_barplot <- Wherb_full %>% group_by(spp, year) %>% summarise(present = sum(pa), N = n()) %>% mutate( percent = present/N*100)
# create plot
H_PA <- ggplot(data=PA_barplot, aes(x=year, y=percent, fill=spp)) +
  geom_bar(stat="identity", position = position_dodge()) + 
  scale_fill_manual(values = c("#cc79a7", "#f0e442"), name = "Species",
                    labels=c("alle" = expression(italic("C. allenii")),
                             "vill" = expression(italic("C. villosissimus")))) +
  ggtitle("Field surveyed herbivory") + 
  ylim(0, 100) +
  xlab("year") +
  theme(plot.title = element_text(hjust = 0.5), legend.text.align = 0) +
  ylab('% of individuals with herbivory')
# print plot
H_PA

# # save plot
# ggsave("wild_herbivory_plot.png", H_PA, device = "png", width = 10, height = 7, units = "cm")
```

We see that both species and year are significant predictors of the presence of herbivory at a \< 0.00001 level, with greater herbivory on *C.allenii* and during the wet year, 2022.

### Percent herbivory (subset)

Here we compare the amount of herbivory as a percentage of leaf area between species for a random subset of wild plants in 2019 and 2022. First, we plot the data.

```{r}
# Read in the 2019 data (refresh)
Wherb19 <- read.csv("vill-all_RLB-herbivory_data_2019-Herbivory.csv", header=TRUE)

# select random subset for herbivory percent analysis
set.seed(1) # use 1 to match paper
Wherb19_subset <- Wherb19 %>% 
  group_by(species) %>% 
  sample_n(10)

# # Non-parametric test of 2019 only
# wilcox.test(Wherb19_subset$per_herbivory ~ Wherb19_subset$species) # W = 90, p-value = 0.0007512
# ## MEANS:  allenii - 2.54; vill - 0.00
# #t.test(Wherb19_subset$per_herbivory ~ Wherb19_subset$species) 
# # t = 3.7367, df = 9, p-value = 0.004649 - differs in signif. use wilcox as appropriate for non-normal data

# Read in the 2022 data 
Wherb22_subset <- read.csv("2022_Wild_Percent_Herbivory_subset.csv", header = T)

# # Non-parametric test of 2022 data only
# wilcox.test(Wherb22_subset$ave_per_herb ~ Wherb22_subset$spp) # W = 96, p-value = 0.0003801
# ## MEANS:  allenii - 3.055  ; vill - 0.1947
# t.test(Wherb22_subset$ave_per_herb ~ Wherb22_subset$spp) 
# # t = 4.873, df = 10.446, p-value = 0.0005718 -> use wilcox as appropriate for non-normal data

# 2019/2022 combined
Wherb19_subset$year <- '2019'
Wherb22_subset$year <- '2022'
names(Wherb19_subset) <- c("date","ID", "spp", "ave_per_herb", "year")
Wherb_full_subset <- Wherb22_subset %>% 
  dplyr::select(spp,ave_per_herb,year) %>% 
  full_join((Wherb19_subset %>% dplyr::select( "spp", "ave_per_herb", "year")),.)

# plotting subset data
# plot of herbivory subset data - probably not for inclusion # 
ggplot(data=Wherb_full_subset, aes(x=year, y=ave_per_herb, fill = spp)) +
  geom_boxplot(outlier.shape=NA) + 
  geom_point(aes(alpha=0.5), position=position_jitterdodge()) +
  scale_fill_manual(values = c("#cc79a7", "#f0e442"), name = "Species",
                    labels=c("alle" = expression(italic("C. allenii")),
                             "vill" = expression(italic("C. villosissimus")))) +
  theme(plot.title = element_text(hjust = 0.5), legend.text.align = 0) +
  ggtitle("Field herbivory subset") +
  xlab("") +
  ylab("percent herbivory")+
  scale_x_discrete(labels=c("alle" = expression(italic("C. allenii")), 
                            "vill" = expression(italic("C. villosissimus"))))

# ggsave("wild_per_herbivory_plot.png", device = "png", width = 10, height = 8, units = "cm")
```

Here we poke at the data with some stats... Not sure what is best here. All analyses say what is pretty obvious from the plot: percent herbivory clearly varies by species, but likely due to 0s, and does not clearly vary here between years - I think this is an artifact of taking a small subset. Beetles generally seem to eat similar (though highly variable) amounts in each year once they get to a suitable plant - they just find fewer suitable plants in drier years (fewer nice moist rolled leaves).

```{r}
# INCOMPLETE: some analyses... 
# Testing model assumptions
# visual assesssment
lm <- lm(ave_per_herb ~ spp + year, data = Wherb_full_subset)
par(mfrow=c(2,2))
plot(lm) # not great - definitely not a normal Q-Q plot, unequal variance btw alle and vill
# homoscedasticity statistical test
ncvTest(lm) # bad... linear regression feels like a really bad idea. 
summary(lm) # reasonable seeming results... year not popping, but makes sense since it was a weaker effect (though still very strong) in the presence absence data

# two way anova to look at the impact of species and year on herbivory
aov2 <- aov(ave_per_herb ~ spp + year, data = Wherb_full_subset)
summary(aov2) # species significant but not year. 
# 2 non-parametric mean comparisons.
wilcox.test(ave_per_herb ~ spp, data = Wherb_full_subset) #W = 372, p-value = 7.095e-07
wilcox.test(ave_per_herb ~ year, data = Wherb_full_subset) #W = 165, p-value = 0.3185
```

## Leaf toughness

We gathered leaf toughness data from wild plants of both species (18 *C. allenii* and 20 *C. villosissimus*) using a leaf penetrometer, which measures the pressure required to poke a hole in the leaf. Here we evaluate whether there is a difference in leaf toughness between species. 

First, we read in and plot the data:

```{r}
# read in the data
tough <- read.csv("leaf.toughness.csv", header = TRUE)

# Plot leaf toughness 
ggplot(data=tough, aes(x=species, y=ave)) +
  geom_boxplot(outlier.shape=NA, aes(fill = species)) + 
  scale_fill_manual(values = c("#cc79a7", "#f0e442"), name = "Species",
                    labels=c("alle" = expression(italic("C. allenii")),
                             "vill" = expression(italic("C. villosissimus")))) +
  ggtitle("Leaf Toughness") +
  geom_jitter(alpha=0.6, width=0.15) +
  theme(plot.title = element_text(hjust = 0.5), legend.text.align = 0) +
  xlab("") +
  ylab("leaf toughness (g)")+
  scale_x_discrete(labels=c("alle" = expression(italic("C. allenii")), 
                            "vill" = expression(italic("C. villosissimus")))) + 
  theme(axis.text.x = element_text(hjust = 0.6, colour = "black")) +
  geom_text(x='vill', y=198, aes(label = "p<0.001"), size=2.5)

#ggsave("leaf_toughness_plot.png", device = "png", width = 10, height = 7, units = "cm")
```
Next, we evaluate the data, find that it best fits a lognormal distribution, and run both a nonpparametric test and t-test on log transformed data.
```{r}
# subset the data to evaluate alle and vill data distributions separately
alle.tough <- tough[tough$species =="alle",] # subsetting alle toughness data
vill.tough <- tough[tough$species =="vill",] # subsetting vill toughness data
shapiro.test(alle.tough$ave) # not normally distributed (tough leaves spike)
hist(alle.tough$ave, breaks=20)
shapiro.test(vill.tough$ave) # not normally distributed, but sorta close; p-value = 0.0765
hist(vill.tough$ave, breaks=20)

# QQ plots of different fits 
qqp(alle.tough$ave, "norm")
qqp(alle.tough$ave, "lnorm") #best 
shapiro.test(log(alle.tough$ave))
qqp(vill.tough$ave, "norm")
qqp(vill.tough$ave, "lnorm") #best 
shapiro.test(log(vill.tough$ave)) # could be normal (not not normal)

wilcox.test(tough$ave ~ tough$species, paired = F) #W = 294, p-value = 0.000566 
t.test(log(tough$ave) ~ tough$species, paired = F) # t = 3.6241, df = 33.24, p-value = 0.0009574
# both say the same thing... which is better to use in the article? 
```

## Leaf chemistry

We have chemistry data for oxidative capacity and chemical concentrations for each of three compound classes: phenolics, flavonoids, and steroidal saponins (aka steroidal glycosides). Here, we compare chemical cconcentrations and oxidative capacity between species.

First, we read in and plot the data:
```{r}
# read in data and adjust column headers
chem <- read.csv('Full_chem_data.csv', header = T)
names(chem) <-  c("Accession", "Species", "DPPH", "Oxidative.Capacity", 
                  "Concentration.of.Phenolics", "Concentration.of.Flavonols",
                  "Concentration.of.NF.Phenolics", "Concentration.of.Saponins")       
# Plot Chemistry data
# Saponins
 S <- ggplot(chem, aes(x=Species, y=Concentration.of.Saponins)) +
  geom_boxplot(outlier.shape=NA, aes(fill = Species)) + 
  scale_fill_manual(values = c("#cc79a7", "#f0e442"), name = "Species") +
  ggtitle("Saponins") + 
  geom_jitter(alpha=0.6, width=0.15) +
  guides(fill="none") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  xlab("") +
  ylab("Concentration of Saponins") + 
  #coord_cartesian(ylim = c(0, 1)) +
  scale_x_discrete(labels=c("ALLE" = expression(italic("C. allenii")), 
                            "VILL" = expression(italic("C. villosissimus")))) + 
  theme(axis.text.x = element_text(hjust = 0.6, colour = "black")) +
  geom_text(x='VILL', y=0.75, aes(label = "p<0.001"), size=3.5) + # y=0.99 for limited y
  theme(axis.text = element_text(size = 12))     

# Non-flavonoid Phenolics
P <- ggplot(chem, aes(x=Species, y=Concentration.of.NF.Phenolics)) +
  geom_boxplot(outlier.shape=NA, aes(fill = Species)) + 
  scale_fill_manual(values = c("#cc79a7", "#f0e442"), name = "Species") +
  ggtitle("Non-Flavonoid Phenolics") +
  geom_jitter(alpha=0.6, width=0.15) +
  guides(fill="none") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  xlab("") +
  ylab("Concentration of non-Flavonoid Phenolics") + 
  #coord_cartesian(ylim = c(0, 1)) +
  scale_x_discrete(labels=c("ALLE" = expression(italic("C. allenii")), 
                            "VILL" = expression(italic("C. villosissimus")))) + 
  theme(axis.text.x = element_text(hjust = 0.6, colour = "black")) + 
  geom_text(x='VILL', y=0.8, aes(label = "p<0.01"), size=3.5) + # y=0.99 for limited y
  theme(axis.text = element_text(size = 12))   

# Flavonids
Fl <- ggplot(chem, aes(x=Species, y=Concentration.of.Flavonols)) +
  geom_boxplot(outlier.shape=NA, aes(fill = Species)) + 
  scale_fill_manual(values = c("#cc79a7", "#f0e442"), name = "Species",
                    labels=c("ALLE" = expression(italic("C. allenii")),
                             "VILL" = expression(italic("C. villosissimus")))) +
  ggtitle("Flavonoids") +
  geom_jitter(alpha=0.6, width=0.15) +
  theme(plot.title = element_text(hjust = 0.5), legend.text.align = 0) + 
  xlab("") +
  ylab("Concentration of Flavonoids") + 
  #coord_cartesian(ylim = c(0, 0.3)) +
  guides(fill="none") +
  scale_x_discrete(labels=c("ALLE" = expression(italic("C. allenii")), 
                            "VILL" = expression(italic("C. villosissimus")))) + 
  theme(axis.text.x = element_text(hjust = 0.6, colour = "black")) + 
  geom_text(x='ALLE', y=0.13, aes(label = "p<0.01"), size=3.5) + # y=0.299 for limited y and x='VILL'
  theme(axis.text = element_text(size = 12))   

# Oxidative capacity 
OC <- ggplot(chem, aes(x=Species, y=Oxidative.Capacity)) +
  geom_boxplot(outlier.shape=NA, aes(fill = Species)) + 
  scale_fill_manual(values = c("#cc79a7", "#f0e442"), name = "Species",
                    labels=c("ALLE" = expression(italic("C. allenii")),
                             "VILL" = expression(italic("C. villosissimus")))) +
  ggtitle("Oxidative Capacity") +
  geom_jitter(alpha=0.6, width=0.15) +
  theme(plot.title = element_text(hjust = 0.5), legend.text.align = 0) + 
  xlab("") +
  ylab("Oxidative Capacity (1-DPPH)") + 
  #coord_cartesian(ylim = c(0, 2.1)) +
  guides(fill="none") +
  scale_x_discrete(labels=c("ALLE" = expression(italic("C. allenii")), 
                            "VILL" = expression(italic("C. villosissimus")))) + 
  theme(axis.text.x = element_text(hjust = 0.6, colour = "black")) + 
  geom_text(x='VILL', y=2.0, aes(label = "p<0.01"), size=3.5) + # y=2.099 for limited y
  theme(axis.text = element_text(size = 12)) 

# Joining the plots with patchwork
 P + S + OC + Fl + plot_annotation(tag_levels = 'A', tag_suffix = ')') & 
  theme(plot.tag.position = c(0, 1),
        plot.tag = element_text(size = 11.5, hjust = 0, vjust = 0))

##ggsave("Chemistry/Chemistry_plot", device = "png", width = 24, height = 20, units = "cm")
#ggsave("Chemistry_plot_free_scale.png", device = "png", width = 18, height = 20, units = "cm")
```
Next, we coonduct t-tests comparing values between species and correct for multiple tests. 
```{r}
# assessing normality
# dividing the data by speceis to assess within species normality
# within group normality is the assumption of an unpaired students/welches t-test
alle_chem <- chem %>% filter(Species=="ALLE")
vill_chem <- chem %>% filter(Species=="VILL")

# # non-Flavonoid Phenolics 
# shapiro.test(alle_chem$Concentration.of.NF.Phenolics) # W = 0.9032, p-value = 0.3508
# qqp(alle_chem$Concentration.of.NF.Phenolics, "norm") 
# shapiro.test(vill_chem$Concentration.of.NF.Phenolics) # W = 0.94897, p-value = 0.7204
# qqp(vill_chem$Concentration.of.Phenolics_nf, "norm") 
# 
# # Flavonoids
# shapiro.test(alle_chem$Concentration.of.Flavonols) # W = 0.96981, p-value = 0.8971
# qqp(alle_chem$Concentration.of.Flavonols, "norm") 
# shapiro.test(vill_chem$Concentration.of.Flavonols) # W = 0.90093, p-value = 0.3367
# qqp(vill_chem$Concentration.of.Flavonols, "norm") 
# 
# # Saponins 
# shapiro.test(alle_chem$Concentration.of.Saponins) # W = 0.86689, p-value = 0.1743
# qqp(alle_chem$Concentration.of.Saponins, "norm") 
# shapiro.test(vill_chem$Concentration.of.Saponins) # W = 0.85908, p-value = 0.1486
# qqp(vill_chem$Concentration.of.Saponins, "norm") 
# 
# # Oxidative Capacity
# shapiro.test(alle_chem$Oxidative.Capacity) # W = 0.74073, p-value = 0.01014
# qqp(alle_chem$Oxidative.Capacity, "norm") # NOT normal
# hist(alle_chem$Oxidative.Capacity, breaks = 20)
# shapiro.test(vill_chem$Oxidative.Capacity) # W = 0.86363, p-value = 0.1631
# qqp(vill_chem$Oxidative.Capacity, "norm") 
# hist(vill_chem$Oxidative.Capacity, breaks = 20)

# Stats
# Phenolics
t.test(chem$Concentration.of.NF.Phenolics~chem$Species, alternative = "two.sided")
# wilcox.test(chem$Concentration.of.Phenolics_nf~chem$Species) # qualitatively equvalent to t.test
# t = 5.0247, df = 6.4714, p-value = 0.001919

# Flavonids
t.test(chem$Concentration.of.Flavonols~chem$Species)
# wilcox.test(chem$Concentration.of.Flavonoids~chem$Species) # qualitatively equvalent to t.test
# t = -3.9658, df = 7.7076, p-value = 0.004465

# Saponins
t.test(chem$Concentration.of.Saponins~chem$Species)
# wilcox.test(chem$Concentration.of.Saponins~chem$Species) # qualitatively equvalent to t.test
# t = 5.5942, df = 11.886, p-value = 0.0001214

# Oxidative activity
t.test(chem$Oxidative.Capacity~chem$Species)
# wilcox.test(chem$Oxidative.Capacity~chem$Species) # W = 49, p-value = 0.0005828 ; stronger result, but both <0.01. 
# t = 4.5951, df = 6.8692, p-value = 0.002624

# multiple test correction
p <- c(0.001919, 0.004465, 0.0001214, 0.002624)
p.adjust(p, method = "BH")
# 0.003498667 0.004465000 0.000485600 0.003498667
```
We see that even with test correction for the four t-tests, chemical concentrations and oxidative capacity are all statistically different between species (p<0.01). 

## Beetle feeding trials
Palatability trials present beetles with a single plant species and measure how much they eat. Choice trials present beetles with both plant species simultaneously and measure how much they eat of each.
 In June 2019, and again in May and June 2022, we collected wild beetles from various species of Costus growing along creeks in the areas surrounding Pipeline road and recorded the species on which beetles were found. To ensure trial beetles weren’t biased towards one of our focal species, we did not collect beetles from either C. villosissimus or C. allenii. Beetles did not have access to leaves for 12 hours before we placed them in either palatability or choice trials. Each beetle was only used in a singel trial before release. To quantify herbivory, we laid a transparency printed with a mm2 grid over the leaf squares and counted mm2 of herbivore damage. 
 
### Palatability
 For palatability trials, we placed a single beetle in a ramekin with a 1.5 cm2 square of either C. villosissimus or C. allenii leaf tissue and quantified the leaf area eaten after 12 hours. In 2019, we conducted 24 palatability trials for each plant species.

First, we import, clean, and plot the palatibility data:
```{r}
# importing beetle trial data
btd <- read.csv("beetle_trials.csv", header = TRUE) 

## palatability trial data
alleBT <- btd[btd$Species.in.trial =="alle",] # subsetting rows with alle as trial
villBT <- btd[btd$Species.in.trial =="vill",] # subsetting rows with vill as trial
alleBT$alle.mm2eaten <- as.numeric(as.character(alleBT$alle.mm2eaten))
villBT$vill.mm2eaten <- as.numeric(as.character(villBT$vill.mm2eaten))
# Reshaping the dataset
singleSPalle <- data.frame(alleBT$Species.found.on, alleBT$Species.in.trial, alleBT$alle.mm2eaten, alleBT$date)
names(singleSPalle) <- c("species.found.on", "trial.species", "mm2.eaten", "date")
singleSPvill <- data.frame(villBT$Species.found.on, villBT$Species.in.trial, villBT$vill.mm2eaten, villBT$date)
names(singleSPvill) <- c("species.found.on", "trial.species", "mm2.eaten", "date")
singleSP <- rbind(singleSPalle, singleSPvill)
# removing 0s (8 from alle, 9 from vill) (common in insect behavior to get 0s that are thrown out unless different between treatments)
singleSP.no0 <- singleSP[singleSP$mm2.eaten != 0,]

#simple box plot of palatability data, shows potential for slight trend towards more vill herbivory.
palatability <- ggplot(singleSP.no0, aes(trial.species, mm2.eaten)) + 
  geom_boxplot(outlier.shape=NA, aes(fill = trial.species)) + 
  ggtitle("Palatability Trials") + 
  geom_jitter(alpha=0.6, width=0.15) +
  scale_fill_manual(values = c("#cc79a7", "#f0e442"), name = "Species") +
  guides(fill="none") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab('herbivory ('~mm^2*')') + 
  xlab("") +
  coord_cartesian(ylim = c(1, 43)) +
  scale_x_discrete(labels=c("alle" = expression(italic("C. allenii")), 
                            "vill" = expression(italic("C. villosissimus")))) + 
  theme(axis.text.x = element_text(hjust = 0.6, colour = "black")) +
  theme(axis.text = element_text(size = 12))   
palatability
```

Next, we evaluate whether beetles find *C. villosissimus* or *C. allenii* more palatable:
```{r}
# check out the distribution of the data:
hist(singleSP.no0$mm2.eaten, breaks = 50) # clearly not normal

# looking at the fit of different distributions using QQ plots
par(mfrow=c(3,1))
qqp(singleSP.no0$mm2.eaten, "norm")
qqp(singleSP.no0$mm2.eaten, "lnorm") # best fit 
MASS::fitdistr(singleSP.no0$mm2.eaten,"Poisson")
qqPlot(singleSP.no0$mm2.eaten, distribution = "pois",lambda = 2) # okay fit 

# regression model - log transformed response and date as a random effect
gBPT  <- lmer(log(singleSP.no0$mm2.eaten) ~ trial.species * (1 | date), data = singleSP.no0)  
summary(gBPT)
# regression model - log transformed response and date as a fixed effect
gBPT2  <- lm(log(singleSP.no0$mm2.eaten) ~ trial.species * date, data = singleSP.no0)  
summary(gBPT2)

# HOW DO I determine if I can leave out date as a random effect and just run a t-test? 

t.test(log(singleSP.no0$mm2.eaten) ~ trial.species, data = singleSP.no0) # t = -1.1442, df = 28.831, p-value = 0.2619

# # regression model - poisson 
# pBPT  <- glmer(mm2.eaten ~ trial.species + (1 | date), 
#                data = singleSP.no0, 
#                family = poisson)  
# summary(pBPT)

# evaluating model assumptions
# compare residuals
# log transformed
qqnorm(resid(gBPT)) 
qqline(resid(gBPT))
# # poisson
# qqnorm(resid(pBPT)) 
# qqline(resid(pBPT))
# # log looks a bit better

# looking for over/under dispersion
# log transformed
gsimOut <- simulateResiduals(gBPT)
plot(gsimOut) # looks good
testDispersion(gsimOut) # p-value = 0.928
# # poisson
# psimOut <- simulateResiduals(pBPT)
# plot(psimOut) # no red flags but doesn't fit as well as log transformed
# testDispersion(psimOut) # p-value = 0.152
# # both don't set off warnings, but log transformed looks better.
```

### Choice
For choice trials, we placed beetles in ramekins (again with wet paper and perforated lids) with one 1.5 cm2 leaf square of both C. villosissimus and C. allenii and quantified the leaf area eaten for each species after 12 hours. We conducted 13 successful choice trials in 2019 and 22 in 2022 (a trial was considered unsuccessful and removed if the beetle did not eat either leaf square). 

Again, we first read in, clean, and plot the data. We will plot it alongside the palatability data for easy comparison.
```{r}
# read in the data
btd <- read.csv("beetle_trials.csv", header = TRUE) 
# subset rows with alle and vill together as trial - aka the choice trials
BCT <- btd[btd$Species.in.trial =="alle/vill",] 
# converting character to numeric
BCT$vill.mm2eaten <- as.numeric(BCT$vill.mm2eaten)
BCT$alle.mm2eaten <- as.numeric(BCT$alle.mm2eaten) 
# remove trials in which beetles did not eat either spp
BCT.no.0 <- BCT[,1:6] %>% filter(!(vill.mm2eaten == 0 & alle.mm2eaten == 0))
# remove trials in which beetles were collected from alle (focal species)
BCT.no.0 <- BCT.no.0 %>% filter(!(Species.found.on == "alle"))

# plotting
BCT.no.0$beetleID <- 1:nrow(BCT.no.0) # assign each trial (each beetle) an ID before splitting each trial into two rows (one for each species) 
# convert to long format
BCT.no.0.long <- BCT.no.0 %>% 
  pivot_longer(cols = c(vill.mm2eaten, alle.mm2eaten),
               names_to = "trial.spp",
               values_to = "mm2eaten")
# remove '.mm2eaten' from species column
BCT.no.0.long$trial.spp <- substr(BCT.no.0.long$trial.spp, 1, 4) 
# box plot of choice data 
choice <- ggplot(BCT.no.0.long, aes(trial.spp, mm2eaten)) + 
  geom_boxplot(outlier.shape=NA, aes(fill = trial.spp)) + 
  scale_fill_manual(values = c("#cc79a7", "#f0e442"), name = "Species",
                    labels=c("alle" = expression(italic("C. allenii")),
                             "vill" = expression(italic("C. villosissimus")))) +
  ggtitle("Choice Trials") + 
  geom_jitter(alpha=0.6, width=0.15) +
  theme(plot.title = element_text(hjust = 0.5), legend.text.align = 0) + 
  xlab("") +
  ylab(bquote("herbivore damage ("~mm^2~"eaten)")) + 
  guides(fill="none") +
  scale_x_discrete(labels=c("alle" = expression(italic("C. allenii")), 
                            "vill" = expression(italic("C. villosissimus")))) + 
  theme(axis.text.x = element_text(hjust = 0.6, colour = "black")) + 
  #geom_text(x='alle', y=33, aes(label = "p<0.05"), size=3.5) +
  theme(axis.text = element_text(size = 12))   

# plot palatability and choice feeding trials together
palatability + choice + plot_annotation(tag_levels = 'A', tag_suffix = ')') & 
  theme(plot.tag.position = c(0, 1),
        plot.tag = element_text(size = 11.5, hjust = 0, vjust = 0))

#ggsave("RLB_feeding_trials_plot", device = "png", width = 13, height = 7, units = "cm")
#ggsave("beetle_choice_plot", device = "png", width = 5, height = 5, units = "cm")
#ggsave("beetle_choice_plot2", device = "png", width = 9, height = 6, units = "cm")

```

Now for the analysis! We first take the difference of mm2 eaten of each species for each beetle/trial to get a single preference value. We then evaluate the impact of a random effect, date, on that preference value. 
```{r}
# take difference of mm2 eaten of each species for each beetle/trial to get a single preference value. (mm2 vill eaten - mm2 alle eaten)
BCT.no.0 <- BCT.no.0 %>% mutate(preference = vill.mm2eaten - alle.mm2eaten)
# assess data distribution
hist(BCT.no.0$preference, breaks=20) # looks normal with gaps haha, so not normal but likely from a normal distribution... I think. 

# evaluate random effect of date on which trial was conducted (temp, humidity, etc of a particular day may impact beetle behavior)
# as fixed effect
BCT_lm <- lm(preference ~ date, data = (BCT.no.0))
summary(BCT_lm) # looks like date doesn't matter
# as random effect (just looking at variance)
BCT_lmer <- lmer(preference ~ (1|date), data = (BCT.no.0))
summary(BCT_lmer) # variance explained by date 2 orders of magnitude lower than residual variance :)
```
We see that, when treated as a fixed effect, no date has a significant (to the .05 level) impact on beetle preference. When treated as a random effect, the variance explained by date 2 orders of magnitude lower than residual variance. Together, this indicates date is not important to include when assessing preference. 
  So, next we compare the mean amount eaten of each species within a trial with a paired t-test to evaluate whether beetles prefer one species over ther other in controlled trial conditions
```{r}
# now that I am convinced the random variable is not strongly influencing preference, I can compare amount eaten of each species within trial with a paired T-test. 
# paired t-test
t.test(BCT.no.0$vill.mm2eaten, BCT.no.0$alle.mm2eaten, paired = T) # t = 1.0992, df = 34, p-value = 0.2794
```
OKAY! So I think this way to analyze the data makes sense, but I need help figuring out how to talk about it. The p-value is definitely not 'significant' (0.28), but there is a trend in the data towards greater herbivory on vill than alle (as seen in the figure above). I also saw this in the other ways I analyzed the data. For example, in a bayesian framework, 80% or so of the posterior distribution supported beetles eating more vill than alle. How do I talk about this in the article to accurately represent the data?

Here is my best guess:
  Roleld leaf beetles, on average, ate more *C. villosissimus* than *C. allenii*. These trends were not significant at the 0.05 level, likely due to high variability in beetle behavior. However, the trend is present in both palatibility and choice trials, and aligns with chemistry and toughness data: beetles are expected to prefer the species with thinner, less chemically defended leaves. 
  This pattern is also contradictory to the clearly greater beetle herbivory on *C. allenii* than *C. villosissimus* seen in the wild. To understand whether this contradiction is due to differing environmental conditions in the habitat of each species, we placed 'herbivory arrays' into intermediate habitat.
  
## Herbivory arrays
  Herbivory arrays consisted of a potted plant of each species (*C. villosissimus* and *C. allenii*) placed near a wild plant of a third species (*C. scaber*) that was colonized by at least one roleld leaf beetle. Arrays were placed in intermediate habitat along a creek on the seasonally dry side of the panamanian isthmus. It was not pooossible to accurately quantify the area eaten (area og beetle damage) after plants were in the field for 3 weeks, so we instead evaluate the proportion of healthy stems that experienced herbivory. (analysis of best guess estimates of area yeild qualitatively the same result)
  
Data importing, cleaning, and plotting:
```{r}
array <- read.csv("Herbivory_arrays_220703.csv", header = T)
names(array) <- c("ID", "array_num", "alt_ID","spp", "site", "date", "herbivory_yn", 
                  "num_RLs", "num_stms_herb", "lf1_herb", "lf1_area", "lf2_herb", "lf2_area", 
                  "lf3_herb", "lf3_area", "lf4_herb", "lf4_area", "lf5.herb", "lf5_area", 
                  "lf6.herbivory", "lf6.area", "lf7.herbivory", "lf7.area", "lf8.herbivory", 
                  "lf8.area", "lf9.herbivory", "lf9.area", "lf10.herbivory", "lf10.area", 
                  "lf11.herbivory", "lf11.area", "notes")

# Filter to only include census dates, select RL count cols, calculate proportion of rls with herbivory
array <- array %>% 
  dplyr::select(ID, array_num, spp, site, date, herbivory_yn, num_RLs, num_stms_herb) %>% # choose only certain columns to include
  filter(!is.na(num_stms_herb)) %>% # remove rows with no data in the 'num_stms_herb' column
  mutate(prop_rl_herb = num_stms_herb/num_RLs)

# plot
ggplot(array, aes(spp, prop_rl_herb)) + 
  geom_boxplot(aes(fill=spp)) + 
  ggtitle("Herbivory Arrays") +
  geom_jitter(alpha=0.6, width=0.15) +
  scale_fill_manual(values = c("#cc79a7", "#f0e442"), name = "Species",
                    labels=c("alle" = expression(italic("C. allenii")),
                             "vill" = expression(italic("C. villosissimus")))) +
  theme(plot.title = element_text(hjust = 0.5), legend.text.align = 0) + 
  xlab("") +
  guides(fill="none") +
  ylab("Proportion of stems with herbivory") + 
  scale_x_discrete(labels=c("alle" = expression(italic("C. allenii")), 
                            "vill" = expression(italic("C. villosissimus")))) + 
  theme(axis.text.x = element_text(angle = 10, hjust = 0.6, colour = "black"))

# ggsave("arrays_lf_proportion_plot.png", device = "png", width = 6, height = 8, units = "cm")
```
Analyses:
```{r}
# regression model - beta distribution (proportion)
# add 0.00001 to 0 values and subtract it from 1 values to fit beta regression
array <- array %>% mutate(prop_rl_herb = case_when(
  prop_rl_herb == 0 ~ 0.00001,
  prop_rl_herb == 1 ~ 0.9999,
  TRUE ~ prop_rl_herb
))

# histogram of response
hist(array$prop_rl_herb, breaks=10)

# regression model
AH <- glmmTMB(prop_rl_herb ~ spp + (1 | site) + (1 | date),
              data = array, 
              family=beta_family(link = "logit"))
summary(AH)  

# evaluating model assumptions
# check residuals
qqnorm(resid(AH)) 
qqline(resid(AH)) # bad
# looking for over/under dispersion
simOut <- simulateResiduals(AH)
plot(simOut) # BAD
testDispersion(simOut) # 
# regardless of model distribution, both random effects (site and date)
# explain almost no variation. The model also does not fit well 
# due to the odd distribution of the response variable,
# Attempting t-test below given that random variables appear unimportant 
# and the difference of amount eaten between species is roughly normal. 

# get proportion of rolled leaves in long format for paired t-test
dat <- array %>% pivot_wider(., id_cols=array_num, names_from = spp, values_from = prop_rl_herb)
# assessing whether difference is normally distributed for use of paired t-test
shapiro.test(dat$alle-dat$vill) # p-value = 0.0359; close to normal
hist(dat$alle-dat$vill)
qqp(dat$alle-dat$vill, "norm")

# paired t-test
wilcox.test(dat$alle, dat$vill, paired = T) # V = 110, p-value = 0.111
t.test(dat$alle, dat$vill, paired = T) # t = 1.6954, df = 25, p-value = 0.1024

# all results are qualitatively (and close to quantitatively) the same: no significant difference between species.
## NOTE - pick up in paper - explain as above - report t-test
```
